\plaintitle

### 2016-09-13 Paper/proceedings.tex
Skeleton version for paper swap.

\documentclass{sigchi}

@@

One of the motivating examples for the power of information visualization is Anscombe's quartet\cite{anscombe1973graphs}, a set of four bivariate sets with nearly identical summary statistics, but with drastically different patterns when drawn as a scatterplot. Implicit in this example is that people have a robust ability to recognize visual patterns in information visualizations and, moreover, to employ this pattern recognition ability towards statistical ends. That is, an underlying assumption in information visualization is that people have the ability to \emph{visually} estimate \emph{statistical} quantities of interest, and can perform at least rough analogs of statistical tasks including model selection, identification of outliers, and estimation of summary statistics, entirely through visual inspection. Knowing the limits of this ``estimation by eye,'' is important for designers of information visualizations seeking to communicate statistical quantities, especially to the general audience.

In this work, we examine

@@

@@

In order to assess the ability of viewers of visualizations to perform regression-by-eye, we conducted a series of three crowd-sourced experiments on Amazon's Mechanical Turk platform. We limited the participant pool to subjects from within the United States, with a prior task approval rating of at least 90\%. For all experiments, we included stimuli where the trend line was visible as attention checks and for validation. We excluded participants with poor performance on these stimuli. Based on timings from internal piloting, we paid each participant \$2 for their participation, for an intended rate of \$8/hour.

We generated the scatterplots by first generating the horizontal position of points through banded Monte Carlo sampling, ensuring that, while the distribution of points in X appeared uniform, at least X/b points appear in each of the b bands. We performed rejection sampling to ensure that no two points overlapped by more than half their radius in the resulting 600x1050 image. Next, we created a set of residual values, sampled evenly from discretized gaussian. This procedure ensured that the actual and intended bandwidth and standard deviation of residual values is precise. This set was permuted, and then applied to the points along with a target linear trend. Heteroskedasticity introduced through permutation could alter the actual linear trend away from the target trend, and so, we performed rejection sampling to ensure that the trend of the resulting points were within $10^{-7}$ of the target trend. We reused this residuals across all different trend types (linear, quadratic, or trigonometric). Except where noted, we selected trend lines that were centered in the image: that is, for a vertical and horizontal data extent of $[0,1]$, $f(0.5) = 0.5$.

@@

\section{Estimation of Slope is Reliable \\ Across Visualizations}

We presented participants with a series of scatterplots, and adjusted a slider to fit the perceived trend in the points. Scatterplots consisted of one of three types of trend (linear, quadratic, or trigonometric). For each scatterplot, participants adjusted a slider that controlled the slope of a rendered trend line (or, in the case of the quadratic trends, the curvature, or the trigonometric, the positive/negative amplitude).

@@

\section{Estimation of Intercept is Subject to \\ ``Within The Bar'' Bias}
\subsection{Hypotheses}
\subsection{Results}
\subsection{Discussion}

\section{Estimation of Slope is Sensitive to Outliers}
\subsection{Hypotheses}
\subsection{Results}
\subsection{Discussion}

>>>>

\documentclass[chi_draft]{sigchi}
%\documentclass{sigchi}

@@

%motivation here is poor. needs another turn of the crank.

\todo{This motivation needs work. Also some redundancies}

One of the motivating examples for the power of information visualization is Anscombe's quartet\cite{anscombe1973graphs}, a set of four bivariate sets with nearly identical summary statistics, but with drastically different patterns when drawn as a scatterplot. Implicit in this example is that people have a robust ability to recognize visual patterns in information visualizations and, moreover, to employ this pattern recognition ability towards statistical ends. That is, an underlying assumption in information visualization is that people have the ability to \emph{visually} estimate \emph{statistical} quantities of interest, and can perform at least rough analogs of statistical tasks including model selection, identification of outliers, and estimation of summary statistics, entirely through visual inspection. Knowing the limits of this ``estimation by eye,'' is important for designers of information visualizations seeking to communicate statistical quantities, especially to the general audience. More than a mere mismatch between the statistical values which the designer intends to communicate, and the ability of viewers to estimate those values, there is also the possibility of \emph{bias}--- the design of visualizations may cause viewers to over- or underestimate task-relevant values.

The estimation of trends in bivariate data is an important analytical task, as it is the basis for many factors relevant to decision-making, such as prediction, imputation, and comparison. Unfortunately, model information is not always explicitly provided by designers of visualizations. When designers do include trend information (for instance, by annotating a scatterplot with a line of best fit), other statistics relevant to the model (such as $r$ values or confidence bands) may be absent. Even if designers specifically include all relevant information, the audience may lack the statistical expertise to interpret these values. As a further complication, visual presentation of bivariate data can influence the types of analysis made by viewers (for instance, viewers may be more likely to consider trends with line charts, and to compare individual values with bar charts \cite{zacks1999bars}). These design choices can also introduce bias (for instance, the visual asymmetry of bars can cause ``within-the-bar'' bias \cite{newman2012bar}). Designers need guidance both on how much accuracy they can expect from viewers making trend estimations ``by eye,'' as well as whether or not different types of bivariate visualizations can bias these estimations.

In this work, we describe a series of crowd-sourced studies on the visual estimation of trend in common bivariate visualizations (scatterplots, area charts, and linegraphs). We report on the results of three studies, investigating estimation of trend slope, trend intercept, and the effect of outliers. We find that, while in most cases viewers make accurate estimations of trends, area charts introduce systematic underestimation of trend intercept, and outliers introduce ambiguity in how trends are estimated. These results suggest that there are several areas where human judgments diverge from the fitted models generated by techniques such as Ordinary Least Squares (OLS), and that the design of bivariate visualizations can introduce additional biases in these judgments.

@@

While there is a great deal of foundational work in visualization and graphical perception dealing with the estimation of individual values in visualizations (such as the height of the bar in a bar chart, or the angle of a line in a linegraph), there is comparatively little work on how viewers of visualizations perceive aggregate statistical quantities. Ariely~\cite{ariely2001seeing} suggests that, in concert with the perception of individual objects, we also collect information about the \emph{ensemble} properties of visual displays. Szafir et al.~\cite{szafir2016four} suggest that this \emph{ensemble coding} can afford the accurate estimation of \emph{summary statistics} in visualizations. However, visualizations with good performance for summary tasks may not result in good performance at point tasks, and vice versa~\cite{albers2014task,fuchs2013evaluation}.

Scatterplots are a standard design for visualization bivariate data, with a multitude of design parameters that affect their suitability for aggregate tasks~\cite{cleveland1984many}. Prior work has confirmed that viewers can make use of scatterplots to perform prediction tasks (which tacitly rely on estimation of trend) in ways that are robust to both noise~\cite{harvey1997effects} and problem frame~\cite{lewandowsky2011popular}. However, the heuristics used when performing these prediction tasks (such as the anchor-and-adjust method~\cite{bolger1993context}) can introduce biases. Similarly, the design of scatterplots (for instance, the decision to encode class membership with color or with shape) can also impact performance at aggregate tasks~\cite{gleicher2013perception,lewandowsky1989discriminating}. The aspect ratio of graphs can also bias judgments about trend~\cite{beattie2002impact}--- wider aspect ratios can cause viewers to underestimate the severity of effects, when compared to narrower charts. Another bias in prediction tasks is the ``within-the-bar'' bias \cite{newman2012bar}: for visually asymmetric visualizations such as bar charts, points contained within the visual area of the bar glyph are perceived as likelier than those outside the glyph.

Recent work in the visualization community has focused on the perception of correlation in scatterplots. Rensink et al.~\cite{rensink2010perception} show that viewers can estimate correlation with some accuracy in scatterplots. Harrison et al. \cite{harrison2014ranking} extend this finding to other visualization types, although a reanalysis by Kay \& Heer~\cite{kay2016beyond} indicates that, for many of the more esoteric bivariate visualizations, performance at this task is poor. Estimation of correlation can also be biased by the design of scales~\cite{cleveland1982variables}--- the whitespace and aspect ratio changes introduced by expanding the scale of the axes can cause overestimation of correlation.

Our task of trend estimation combines elements of both prediction and correlation estimation tasks. As with prediction, there is not necessarily an unambiguous ``correct'' estimation of trend (different modeling and regression methods can produce different trend lines). As with correlation, viewers must make holistic judgments about the dataset in a way that (as per Harrison et al.) likely relies on a set of key visual proxies. And, as with both tasks, we believe that trend estimation can be biased through conscious or unconscious design choices.

@@

In order to assess the ability of viewers of visualizations to estimate trend, we conducted a series of three crowd-sourced experiments on Amazon's Mechanical Turk platform. Crowd-sourcing is a valuable tool for graphical perception experiments, as it affords flexibilities and scales that would be difficult for in-person experiments~\cite{heer2010crowdsourcing}. Crowd-sourced graphical perception experiments produce results that are largely in keeping with prior, lab-based work~\cite{talbot2014four}.

We limited the participant pool to subjects from within the United States, with a prior task approval rating of at least 90\%. For all experiments, we included stimuli where the trend line was visible as attention checks and for validation. We excluded participants with poor performance on these stimuli. Based on timings from internal piloting, we paid each participant \$2 for their participation, for an intended rate of \$8/hour.

We generated the scatterplots by first generating the horizontal position of points through banded Monte Carlo sampling, ensuring that, while the distribution of points in $X$ appeared uniform, at least $X/b$ points appear in each of the $b$ bands. We performed rejection sampling to ensure that no two points overlapped by more than half their radius in the resulting 600x1050 image. Next, we created a set of residual values, sampled evenly from discretized gaussian. This procedure ensured that the actual and intended bandwidth and standard deviation of residual values is precise. This set was permuted, and then applied to the points along with a target linear trend. As heteroskedasticity introduced through permutation could alter the actual linear trend away from the target trend, we performed rejection sampling to ensure that the trend of the resulting points were within $10^{-7}$ of the target trend. We reused this residuals across all different trend types (linear, quadratic, or trigonometric). Except where noted, we selected trend lines that were centered in the image: that is, for a horizontal data extent of $[0,1]$, $f(0.5) = 0.5$.

\todo{Add figure of experimental apparatus}

In all experiments, participants entered in information using a slider without tick marks, in order to avoid anchoring effects \cite{matejka2016effect}. Moving the slider adjusts a purple trend line by adjusting its slope (in experiments 1 and 3) or its y-intercept (in experiment 2). After moving the slider, participants had to confirm their choice of trend line. For validation, certain trials contained the actual (OLS) trend line. We excluded the results of participants with poor performance on these validation questions.

@@

\todo{48 per experiment, report all demographics here rather than in experiment subsections.}

\section{1. Estimation of Slope is Reliable \\ Across Visualizations}

We presented participants with a series of bivariate visualizations, and adjusted a slider to fit the perceived trend in the points. Visualizations were one of a scatterplot, linegraph, or area chart (with the filled area below the line). Visualizations consisted of one of three types of trend (linear, quadratic, or trigonometric). For each stimulus, participants adjusted a slider that controlled the slope of a rendered trend line (or, in the case of the quadratic trends, the curvature, or the trigonometric, the positive/negative amplitude).

@@

\section{2. Estimation of Intercept is Subject to \\ ``Within The Bar'' Bias}

As with the previous experiment, we presented participants with a series of bivariate visualizations. However, instead of estimating the \emph{slope} of the points, participants estimated the \emph{y intercept} of the trend line. For each trial, we added a uniform offset to the points in the bivariate visualization in the data range $[-0.25,0.25]$. The rendered trend line was initially placed with the correct slope, and such that $f(0.5)=0.5$ in data space. Participants adjusted a slider controlling the vertical offset of this trend line.

\section{3. Estimation of Slope is Sensitive to Outliers}

This experiment was largely identical to experiment 1 above, except we designated 0, 5, 10, or 15 points at the very beginning, middle, or end of the series as outliers, and placed them randomly in the top or bottom 10\% of the visual area of the visualization (whichever was furthest from the trend line). We then calculated the outlier-sensitive line of best fit, as well as the intersection between this new trend line and the original, intended trend line. As with experiment 1, the participants controlled the slope of a rendered trend line with a slider. However, rather than being placed such that $f(0.5)=0.5$, we offset the trend line to the intersection of the robust (without outliers) and non-robust (including outliers) trend line. This allowed participants to express both types of fit with the same slider interaction, while affording estimations beyond a simple interpolation of both trends.
